---
title: "Examination of Health Professional Shortage Area in Arizona and New Mexico"
subtitle: "Intro to Data Science: Assignment 04"
author: "Maddie Dimarco - msd150 & Hannah Reynolds - hjr45"
execute: 
  warning: false
format:
  html:
    embed-resources: true
---

## Data

This data is from the Health Resources and Services Administration  <https://data.hrsa.gov/tools/shortage-area/hpsa-find>.

Uploading and cleaning data:
```{r}
#load libraries
library(tidyverse)
library(ggplot2)
library(haven)
library(here)
library(ggthemes)

#load and clean data

data <- read_csv(
  here("Hpsa_Find_Export.csv")
  ) 

data <- data %>%
  filter(PrimaryStateName == "New Mexico" | PrimaryStateName == "Arizona")
```
## Visualization 1: Frequency of HPSA Designation by State

These two bar graphs show the frequency of the three HPSA Designations - dental health, mental health, and primary care - for Arizona and New Mexico. The graphs show which areas of health care experience the greatest shortages in each state. Both Arizona and New Mexico experience the greatest shortages in primary care, however, dental health also greatly lacks health professionals in Arizona. 

```{r}
data %>%
  filter(PrimaryStateName == "Arizona") %>%
  ggplot(aes(x=Discipline)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Frequency of HPSA Designation in Arizona",
    x = "",
    y = "Frequency",
    caption = "Source: HRSA"
  ) +
  theme_economist_white()
  
data %>%
  filter(PrimaryStateName == "New Mexico") %>%
  ggplot(aes(x=Discipline)) +
  geom_bar(position = "dodge") +
  labs(
    title = "Frequency of HPSA Designation in New Mexico",
    x = "",
    y = "Frequency",
    caption = "Source: HRSA"
  ) +
  theme_economist_white()
```

## Visualization 2

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
#| echo: false
2 * 2
```

## Visualization 3

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
#| echo: false
2 * 2
```

## Visualization 4

When you click the **Render** button a document will be generated that includes both content and the output of embedded code. You can embed code like this:

```{r}
#| echo: false
2 * 2
```

The `echo: false` option disables the printing of code (only output is displayed).
